Subglacial and subaerial fluvial sediment transport capacity respond differently to water discharge variations

### 2024-04-05 Delaneyetal2024.tex
Delaneyetal2024.tex

\begin{keypoints}
\item Short water discharge fluctuations in subglacial channels are accommodated by water velocity, not channel size, as in subaerial systems.
\item Greater variability in water velocity causes sediment transport capacity to vary more in subglacial channels than in subaerial channels.
\item Significant subglacial sediment transport capacity can persist across a range of water discharges, impacting the  timing of peak events.
\end{keypoints}

\begin{abstract}
  Sediment transport capacity in both subaerial and subglacial channels depends on the shear stress exerted across the channel bottom, which varies with the water velocity and channel width.
  In subaerial channels, water discharge variations are accommodated  by flow depth and width changes, along with water velocity.
  In subglacial channels rather, water discharge changes primarily affect solely water velocity due to evolving channel geometry over hours to weeks.
  Here, hydrographs from an Alpine glacier and the Greenland Ice Sheet are used to drive subglacial and subaerial hydraulics models, revealing greater variability in the inferred sediment transport capacity in subglacial than in subaerial channels.
  Results also show that high subglacial sediment transport rates and water velocities can occur across wide-ranging water discharges.
  These findings are discussed in the context of understanding sediment discharge from glaciers with different hydrological forcings and assumptions of channel shape.
\end{abstract}

\section*{Plain Language Summary} % 200 words
Sediment export from glaciers impacts downstream ecosystems and communities and has changed in response to climate warming.
Sediment export depends partly on the sediment transport capacity, and the amount of sediment that could be transported for the hydraulic conditions with abundant sediment availability.
This capacity largely depends on the width of the channel and the water's velocity.
In subaerial channels, like rivers, water velocity, and channel width both respond rapidly to water discharge.
However, subglacial channels are pressurized by the ice above, so water flows through a subglacial pipe that evolves in size.
This pressurized flow means that channel size evolves slowly compared to variations in water discharge, therefore these variations are largely accommodated by changes in water velocity as opposed to channel size.
Models of these processes show that sediment transport capacity is highly variable in subglacial channels, and a wide range of water discharges can result in comparable sediment transport capacities.
In subaerial channels, by contrast, sediment transport capacity is less variable, and variations depend solely on water discharge, not the channel size.
The channel's different responses to water discharge may impact the interpretation of sediment transport records from glacierized catchments, complicating the link between glacier melt and sediment discharge.

@@

Over long periods, such as millennia, processes such as glacier abrasion and quarrying sculpt landscapes and create sediment to be transported fluvially from under glaciers \cite<c.f.>{hallet1979,iverson2012,ugelvig2018}.
Pressurized subglacial water can transport this sediment from underneath glaciers \cite{walder1994,creyts2013,beaud2018,delaney2019}, should enough sediment be present subglacially.

@@

The response of water velocity to changing water discharge in subglacial channels differs from subaerial ones.
The size of subglacial channels responds to the creep closure of the ice and the opening of the channel by melting due to frictional heating of water flowing through the channel \cite{rothlisberger1972}.
As a result, the subglacial channel size only evolves relatively slowly over days to months, whereas water discharge can vary over hours \cite<e.g.>{iken1986,andrews2014,nanni2020}.
Therefore, on short time scales, subglacial water flow behaves more like pipe flow, and changes in water discharge $Q$ are mainly accommodated by increased or decreased water velocity $v$ \cite<Equation~\ref{eq:v} and Figure~\ref{fig:cartoon}; >{alley1997}.

@@

Because of the above, sediment mobilization in subaerial and subglacial channels responds differently to changing water discharge.
These differences are implicitly included in a wide range of available models that  quantify sediment transport in both subglacial and subaerial channels  \cite<e.g.>{walder1994,alley1997,tucker1997,creyts2013,delaney2019,hewitt2019,wickert2019}.
The divergent response of changing water discharge on sediment mobilization capacity may well impact sediment dynamics near glacier margins where flow transitions from pressurized to open channel flow \cite<e.g.>{perolo2018}.
Additionally, the variable response of sediment transport capacity to water discharge in the two systems may affect the interpretation of sediment transport records in glacial systems \cite<e.g.>{muller1968,richards2003,swift2005,ganti2016}.
Yet, little work has examined the differing relationship between variations in water discharge and sediment discharge in subglacial and subaerial channels \cite{alley1997}.

This manuscript evaluates the relationship between sediment transport capacity and water discharge in subglacial channels and compares it to subaerial ones.
Models of subglacial and subaerial hydraulics are used to quantify the respective water velocity and channel morphology.
The models are applied to hydrological records from an Alpine glacier in Switzerland (Fieschergletscher) and  a land-terminating glacier in Greenland (Leverett Glacier).
The  lumped nature of the models isolates the disparity between water discharge and sediment transport capacity in subglacial systems, independent of the upstream drainage network and sediment access.
Outputs demonstrate differences in the relationships between  water discharge, channel geometry, and water velocity in the subglacial and subaerial channels.
The manuscript then uses the demonstrated differences between the two channel types to discuss the implications for interpreting the processes driving variability in sediment transport records and their relationship with hydrology from glacierized catchments.

@@

The Leverett Glacier in Greenland (labeled \icesheet{}) serves as the ice sheet setting.
Water discharge was collected roughly $2$\unit{km} downstream from the terminus between June and August/September 2009 to 2012 ($67^\circ\,03'\,5''$ N, $50^\circ\,12'\,59''$ W).
Data is used in \citeA{tedstone2017} and \citeA{hawkings2014} at a $5$\,\unit{min} time interval from May 28, 2012 to August 8, 2012.

Note that sediment discharge records are omitted from the analysis here given the role of sediment supply, in addition to capacity, on capturing variations in sediment discharge \cite<e.g.>{delaney2024}.

\section{Methods}
\label{sect:meth}
The models described below (Sections ~\ref{sect:sub_mode}~and~\ref{sect:fluv}) represent relationships amongst water discharge, water velocity, and channel geometry in both subaerial and subglacial channels (Table \ref{table:vpm}).
Both models use the measured discharge to calculate water velocity, shear stress, and width-integrated shear stress, upon which both suspended sediment and bedload transport depend \cite<Figure \ref{fig:cartoon}; >{shields1936}.
The evaluation of these variables omits the selection of a sediment transport relationship and a grain-size parameter \cite<e.g.>{shields1936,meyer1948}.

@@

Following the Darcy-Weisbach equation, the head drop $\Delta h$ is
\begin{linenomath*}
  \begin{equation}
    \label{eq:dh}
    \Delta h \,  = l \,\frac{1}{2g} \,f_r\,\frac{v_g^2}{D_h},
  \end{equation}
\end{linenomath*}
\noindent where $f_r$ is a friction factor, $D_h$ is the hydraulic diameter, $l$ is the channel length, and $v_g=\frac{Q}{S_g}$ is the water flow speed.
%
The hydraulic diameter $D_h$ is converted to wetted area $S_g$ with
\begin{equation}
  \label{eq:Dh2S}
  S_g= \frac{D_h^2}{2}\,\frac{(\frac{\beta}{2}+\sin \frac{\beta}{2})^2}{\beta - \sin \beta},
\end{equation}
where $\beta$ is the central angle of the circular segment that comprises the channel (the so-called Hooke angle, \citeA{hooke1990}). Note that $\beta =\pi$ corresponds to a semi-circular channel and smaller values of $\beta$ result in shallow, wide channels.
This completes the subglacial hydraulic model which is described by the state variables $S_g$ and $\Delta h$.

@@

The models above are applied to proglacial discharge records from the Fieschergletscher (scenario \alpine{}) and the Leverett Glacier (scenario \icesheet{}), using subglacial model outputs that are contrasted with subaerial ones.
Outputs of the models represent generalizable sediment transport characteristics from these hydrographs, rather than actual hydraulic conditions.
To generalize these scenarios, \alpine{}  is exemplified by Fieschergletscher's relatively thin ice thickness \cite<$h_{ice}$: $225$\,\unit{m}>{grab2021}, low water discharge ($\sim\,10$\,\unit{m}$^3$\,\unit{s}$^{-1}$) and high diurnal variability in water discharge (Figure~\ref{fig:model_outs}, a).
\icesheet{}  is exemplified by Leverett glacier's  thick ice  \cite<$h_{ice}$: $700$\,\unit{m}; >{morlighem2017}, high water discharge \cite<$\sim\,300$\,\unit{m}$^3$\,\unit{s}$^{-1}$>{tedstone2017}  and low diurnal variability in water discharge (Figure~\ref{fig:model_outs}, e).
%However, both scenarios share a discharge with a similar, relative seasonal amplitude.

@@

\section{Results}
\begin{figure}[h]
  \centering
  \includegraphics[width=0.9\linewidth]{Fig2.pdf}
  \caption{Model outputs from simulations using the hydrographs in panels a and e for the scenarios \alpine{} (a-d) and \icesheet{} (e-h).
    Gray (black) lines in a and e represent the subglacial channel size (water discharge).
    Blue (orange) lines represent outputs from the subglacial (subaerial) channel.
    Data are shown at $15$\unit{min} intervals.
    Arrows denote where variables peak in the subglacial channel before the subaerial one.
    Insets in f-h show the  peak melt event denoted by the shaded area in panels e--h, but have arbitrary y-axis.
    % Note that the x and y axes are different for \alpine{} and \icesheet{}.
  }
  \label{fig:model_outs}
\end{figure}

Following the first experiment defined in Sections~\ref{sect:imp}, model outputs exhibit different seasonal evolutions and peaks.
Water velocity and shear stress represent the potential for sediment mobilization, and width-integrated shear stress represents a proxy for the total sediment transport capacity across the channel bed (Figure~\ref{fig:model_outs}).
Furthermore, the subglacial model outputs are generally larger than their subaerial equivalents, thus greater sediment transport capacity would be expected in subglacial channels compared to subaerial ones.

@@

Examining the relationship between water discharge and the model outputs, a different relationships emerge between the subaerial and subglacial cases, due to the hysteresis in channel size in the subglacial model (Equation~\ref{eq:dS_dt}).
Because there is no history dependence in  subaerial channels, each water discharge value in the subaerial channel results in a unique water velocity, shear stress, and width-integrated shear stress (Section~\ref{sect:sub_mode}).
This characteristic results in a perfect rank correlation between the variables (Figure~\ref{fig:Qw_vari}).

@@

The second numerical experiment aims to evaluate the role of channel shape and friction factors on the findings above and to examine the effect of variability in water discharge.
Across the range of channel shapes and friction factors  examined (Section\,\ref{sect:imp}), variability in water velocity and shear stress remains higher in the subglacial system compared to the subaerial one in both \alpine{} and \icesheet{}.
This suggests that there is greater intermittency and variability in sediment transport in subglacial systems compared to subaerial ones, especially in supply-limited systems where sediment could be exhausted or mobilized over diurnal discharge cycles.
In contrast, in \alpine{}, the variability in width-integrated shear stress is comparable between the subaerial and subglacial systems, due to the highly variable water discharge, largely due to diurnal variations (Figure~\ref{fig:multi_run}).
However, in \icesheet{}, with lower water discharge fluctuations, the variability in width-integrated shear stress remains consistently higher in the subglacial system across the time aggregation periods.

Results from  both \alpine{} and \icesheet{} scenarios show that the variability in subglacial  decreases with aggregations longer than approximately 1--5 days (Figure~\ref{fig:multi_run}).
Yet, substantial differences in the variability between the model outputs for subglacial and subaerial cases persist in periods of up to $15$ days, longer than which seasonal variations might be conflated.
Note that the timing of peak sediment transport conditions still occurs, across these periods, and they are are thus not captured in aggregating water discharge over longer periods (Figure~\ref{fig:model_outs}).

@@

Model outputs (water velocity, shear stress, and width-integrated shear stress) vary differently in response to water discharge in subglacial channels compared to subaerial ones (Figure~\ref{fig:Qw_vari}).
In sediment transport relationships that evaluate sediment discharge capacity, such as in \citeA{meyer1948} or \citeA{engelund1967}, shear stress is scaled to the power of $\frac{3}{2}$ or $\frac{5}{2}$, respectively.
The exponent greater than $1$ magnifies sediment discharge variability beyond the variable sediment transport parameters described above (Figure~\ref{fig:multi_run}; Table~\ref{tab:Qs}).

Sediment transport capacity in subglacial channels responds more rapidly to changes in water discharge compared to their subaerial counterparts \cite<Figures~\ref{fig:Qw_vari}~and~\ref{fig:multi_run}>{alley1997}.
In subglacial R\"othlisberger channels in steady state with water discharge, the width-integrated shear stress response to water discharge is $ w_g\,\tau_{g}\, \propto\,  Q^{\sim\frac{4}{5}}$ (Table\,\ref{tab:eqs1}).
However, subglacial channels rarely operate in a steady state \cite<e.g.>{gimbert2016}.
For sufficiently short periods, channel width and wetted area are fixed in the subglacial system (akin to pipe flow) so that the response of width-integrated shear stress is $w_g\, \tau_{g}\, \propto\,  Q^{2}$ (Table\,\ref{tab:eqs1}).
\ref{sect:scaling} provides a complete description of sediment transport capacity in different drainage conditions.
Note that in pipe flow conditions, maximum sediment transport would coincide with peak water discharge and a co-varying relationship with water discharge would occur, unlike with the evolving channel size here where a disparate relationship with water discharge occurs (Section~\ref{sect:sub_mode}; Figure \ref{fig:model_outs} and \ref{fig:Qw_vari}). 
Similar adjustments in channel shape occur in subaerial channels \cite{phillips2016}, albeit at timescales longer than the subglacial channel's response time of hours to days.

Substantial differences emerge in model outputs between the different hydrological forcings of the \alpine{} and \icesheet{} cases.
Indeed, the more decoupled and sporadic relationship between model outputs and water discharge in \alpine{} results from the
relatively larger discharge variations on daily to weekly timescales compared to \icesheet{}.
This occurs as the evolution of the R-channel's size is small compared to the variations in water discharge.
Such water discharge variability in alpine catchments may cause shear stress to approach the water discharge squared \cite< \ref{sect:scaling}; c.f.>{alley1997}.
With respect to \icesheet{}, a closer approximation of the hydraulics could be represented by pipe flow here, where the channel size remains fixed.
Conversely, the reduced variability in water discharge in \icesheet{} may result in the R-channel's size being closer to equilibrium concerning the variations in water discharge.
In this case, the exponent on water discharge for shear stress is likely less than $2$, but greater than $\frac{4}{5}$ occurring when the R-channel's size evolves in equilibrium with water discharge (see \ref{sect:scaling}).
As a result, a stronger correlation persists between subglacial model outputs and water discharge in \icesheet{} compared to \alpine{} (Figure~\ref{fig:Qw_vari}).

@@

The stronger link between sediment transport capacity and water discharge in subaerial channels may mean that variations in sediment transport records can be more easily attributed to hydro-climatic conditions than in subglacial systems.
A glacier's sediment transport capacity is impacted by the ice thickness controlling the channel closure rate, and the glacier's surface slope, in addition to water discharge and sediment size \cite<Figure~\ref{fig:multi_run}, Section~\ref{sect:sub_mode}; >{rothlisberger1972,shreve1972,gimbert2016,stevens2022,walder1994}.
In a supply-limited state, the glaciers' sediment discharge record also responds to  sediment availability and bedrock erosion from  water pressure variations and sliding  \cite<e.g.>{iverson2012,herman2015,delaney2023}.
This multitude of processes lies in contrast to many subaerial channels, where transport capacity typically responds to sediment size, channel shape, water discharge, and hydraulic gradient.
The hydraulic gradient and channel shape can remain stable over years to millennia \cite<Section~\ref{sect:fluv}; e.g.>{tucker1997,wickert2019}.

@@

Furthermore, the greater variations in subglacial sediment transport capacity compared to subaerial ones could cause a supply-limited regime to persist at many glaciers.
In subglacial systems, sediment's critical shear stress, or threshold at which sediment mobilization occurs, can be reached more frequently and across a range of water discharges, compared to subaerial systems (Figure~\ref{fig:Qw_vari}).
Sediment exhaustion through this process may  explain the stronger dependence of sediment discharge from the Greenland Ice Sheet on basal shear stress, a proxy for bedrock erosion, rather than glacier melt \cite{overeem2017}.

@@

Numerical models of subglacial sediment transport generally apply classical parameterizations of subglacial hydrology to evaluating sediment transport conditions \cite<e.g>{hewitt2019,beaud2018}.
In these R-channel models, realistic water pressure regimes can be achieved by changing the channels shape and friction factor, therefore  model tuning is done with the friction factor and leaving the channel shape fixed as a semi-circle \cite<$\beta =\pi$; >{werder2010b}.
However, the role of channel width in the findings here suggests that more thorough parameterization of channel shape and channel width is needed in assessing subglacial sediment transport capacity as compared to water pressure (Section\,\ref{sect:sub_mode}, \ref{sect:scaling}).

Water discharge measurements at the hourly timescale remain unavailable in many catchments, further limiting the ability of subglacial  sediment transport models to capture certain events.
The model outputs' variability in subglacial channels approaches that of subaerial ones over $1-5$ days of  aggregation (Figure~\ref{fig:multi_run}), possibly resulting in parity in sediment transport capacity variations between subaerial and subglacial channels over these periods.
Note that although variability between the events could remain comparable, water discharge forcings aggregated over periods longer than days may omit key sediment discharge characteristics, such as the timing of peaks in sediment transport capacity from glaciers (Figure~\ref{fig:model_outs}).

@@

Implications of the presented findings include:
\begin{enumerate}
\item the changing relationship between water discharge and sediment transport capacity below glaciers creates larger variability in sediment transport capacity compared to river systems;
  \vspace{.1cm}

\item
  water discharge variability impacts subglacial sediment transport capacity fluctuations to a larger degree than subaerial channels;
  \vspace{.1cm}

\item the uneven response of sediment transport capacity to water discharge variations shows a different response and timing to periods of elevated water discharge in subglacial channels compared to subaerial ones.
However, representing subglacial sediment transport capacity with either open-channel flow or pipe flow neglects differences in timing.
\end{enumerate}

>>>>

\begin{keypoints}
\item Short water discharge fluctuations in subglacial channels are accommodated by water velocity, not channel size as in subaerial systems.
\item Greater variability in water velocity causes sediment transport capacity to vary more in subglacial channels than in subaerial channels.
\item Significant subglacial sediment transport capacity can persist across a range of water discharges, impacting the  timing of peak events.
\end{keypoints}

\begin{abstract}
  Sediment transport capacity in both subaerial and subglacial channels depends on the shear stress exerted across the channel bottom, which varies with water velocity and channel width.
  In subaerial channels, water discharge variations are accommodated  by flow depth and width changes, along with water velocity.
  However, in subglacial channels  water discharge changes mainly result in velocity variations, as the channel geometry evolves slowly (over days).
  Here, hydrographs from an Alpine glacier and the Greenland Ice Sheet are applied to subglacial and subaerial hydraulics models, revealing greater variability in the inferred sediment transport capacity in subglacial than in subaerial channels.
  Results also show that high subglacial sediment transport rates and water velocities can occur across wide-ranging water discharges.
  These findings are discussed in the context of understanding sediment discharge from glaciers with different hydro-climatic forcings and assumptions of channel shape.
\end{abstract}

\section*{Plain Language Summary} % 200 words
Sediment export from glaciers impacts downstream ecosystems and communities and has changed in response to climate warming.
Sediment export depends on two main factors: sediment availability, and the capacity of the hydrological system to transport that sediment.
This capacity largely depends on the width of the channel and the water's velocity.
In subaerial channels, like rivers, water velocity, and channel width both respond rapidly to water discharge.
However, subglacial channels are pressurized by the ice above, so water flows through a subglacial pipe that evolves in size only slowly.
Variations in water discharge and thus also sediment transport capacity are therefore mainly accommodated by changes in the water velocity.
Models of these processes show that sediment transport capacity is highly variable in subglacial channels, and a wide range of water discharges can result in similar sediment transport capacities.
In subaerial channels, by contrast, sediment transport capacity is less variable, and variations depend solely on water discharge, not the channel size.
These findings may impact the interpretation of sediment transport records from glacierized catchments, complicating the link between glacier melt and sediment discharge.

@@

Over long periods, such as millennia, processes such as glacier abrasion and quarrying sculpt landscapes and create sediment to be transported fluvially from under glaciers \cite<c.f.>{hallet1979,iverson2012,ugelvig2018}.
Pressurized subglacial water can transport this sediment from underneath glaciers \cite{walder1994,creyts2013,beaud2018,delaney2019}, should adequate sediment be present subglacially.

@@

The response of water velocity to changing water discharge in subglacial channels differs from subaerial ones.
The size of subglacial channels is controlled on the opposing processes of channel opening by frictional heating of water flow on the one hand versus creep closure by ice flow on the other \cite{rothlisberger1972}.
As a result, the subglacial channel size only evolves relatively slowly over days or longer, whereas water discharge can vary over hours \cite<e.g.>{iken1986,andrews2014,nanni2020}.
Therefore, on short time scales, subglacial water flow behaves more like pipe flow, and changes in water discharge $Q$ are mainly accommodated by increased or decreased water velocity $v$ \cite<Equation~\ref{eq:v} and Figure~\ref{fig:cartoon}; >{alley1997}.

@@

Sediment mobilization in subaerial and subglacial channels, therefore, responds differently to changing water discharge.
These differences are implicitly included in a wide range of available models that  quantify sediment transport in both subglacial and subaerial channels  \cite<e.g.>{walder1994,alley1997,tucker1997,creyts2013,delaney2019,hewitt2019,wickert2019}.
The different response of  sediment transport capacity to changing water discharge  may well impact sediment dynamics near glacier margins where flow transitions from pressurized to open channel flow \cite<e.g.>{perolo2018}.
Additionally, the variable response of sediment transport capacity to water discharge in the two systems may affect the interpretation of sediment transport records in glacial systems \cite<e.g.>{muller1968,richards2003,swift2005,ganti2016}.
Yet, little work has examined the differing relationship between variations in water discharge and sediment discharge in subglacial and subaerial channels \cite{alley1997}.

This manuscript evaluates the relationship between sediment transport capacity and water discharge in subglacial channels and compares it to subaerial ones.
Models of subglacial and subaerial hydraulics are used to quantify the respective water velocity and channel morphology.
The models are applied to proglacial hydrological records from an Alpine glacier in Switzerland (Fieschergletscher) and  a land-terminating glacier in Greenland (Leverett Glacier).
The  lumped nature of the models isolates the relationship between water discharge and sediment transport capacity in subglacial and subaerial systems, independent of the upstream drainage network and sediment access.
Outputs demonstrate differences in the relationships between  water discharge, channel geometry, and water velocity in the subglacial and subaerial channels.
The manuscript then uses the demonstrated differences between the two channel types to discuss the implications for interpreting the processes driving variability in sediment transport records and their relationship with hydrology from glacierized catchments.

@@

The Leverett Glacier in Greenland (labeled \icesheet{}) serves as the ice sheet setting.
Water discharge was measured roughly $2$\unit{km} downstream from the terminus between June and August/September 2009 to 2012 ($67^\circ\,03'\,5''$ N, $50^\circ\,12'\,59''$ W).
Data is used in \citeA{tedstone2013} at a $5$\,\unit{min} time interval from May 28, 2012 to August 8, 2012.

Note that we omit analysis suspended sediment discharge records due the role of sediment supply on capturing variations in sediment discharge, in addition to sediment transport capacity \cite<e.g.>{delaney2024}.

\section{Methods}
\label{sect:meth}
The models described below (Sections ~\ref{sect:sub_mode}~and~\ref{sect:fluv}) represent relationships amongst water discharge, water velocity, and channel geometry in both subaerial and subglacial channels (Table \ref{table:vpm}).
Both models use the measured discharge to calculate water velocity, shear stress, and width-integrated shear stress, upon which both suspended sediment and bedload transport depend \cite<Figure \ref{fig:cartoon}; >{shields1936}.
The evaluation of shear stress omits the selection of a sediment transport relationship and a grain-size parameter \cite<e.g.>{shields1936,meyer1948}.

@@

Following the Darcy-Weisbach equation, the head drop $\Delta h$ is
\begin{linenomath*}
  \begin{equation}
    \label{eq:dh}
    \Delta h \,  = l \,\frac{1}{2g} \,f_r\,\frac{v_{g}^{2}}{D_h},
  \end{equation}
\end{linenomath*}
\noindent where $f_r$ is a friction factor, $D_h$ is the hydraulic diameter, $l$ is the channel length, and $v_g=\frac{Q}{S_g}$ is the water flow speed.
%
The hydraulic diameter $D_h$ is converted to wetted area $S_g$ with
\begin{equation}
  \label{eq:Dh2S}
  S_g= \frac{D_h^2}{2}\,\frac{(\frac{\beta}{2}+\sin \frac{\beta}{2})^2}{\beta - \sin \beta},
\end{equation}
where $\beta$ is the central angle of the circular segment that comprises the channel (the so-called Hooke angle, \citeA{hooke1990}). Note that $\beta =\pi$ corresponds to a semi-circular channel and smaller values of $\beta$ result in shallow, wide channels.
This completes the subglacial hydraulic model which is described by the state variables $S_g$ and $\Delta h$.

@@

The models above are applied to proglacial discharge records from the Fieschergletscher (scenario \alpine{}) and the Leverett Glacier (scenario \icesheet{}).
Outputs of the models represent generalizable sediment transport characteristics from these hydrographs, rather than actual hydraulic conditions.
To generalize these scenarios, \alpine{}  is exemplified by Fieschergletscher's relatively thin ice thickness \cite<$h_{ice}$: $225$\,\unit{m}>{grab2021}, low water discharge ($\sim\,10$\,\unit{m}$^3$\,\unit{s}$^{-1}$) and high diurnal variability in water discharge (Figure~\ref{fig:model_outs}, a).
\icesheet{}  is exemplified by Leverett glacier's  thick ice  \cite<$h_{ice}$: $700$\,\unit{m}; >{morlighem2017}, high water discharge ($\sim\,300$\,\unit{m}$^3$\,\unit{s}$^{-1}$)  and low diurnal variability in water discharge (Figure~\ref{fig:model_outs}, e).
%However, both scenarios share a discharge with a similar, relative seasonal amplitude.

@@

\section{Results}
\begin{figure}[h]
  \centering
  \includegraphics[width=0.9\linewidth]{Fig2.pdf}
  \caption{Model outputs from simulations using the hydrographs in panels a and e for the scenarios \alpine{} (a-d) and \icesheet{} (e-h).
    Black (gray) lines in a and e represent the subglacial channel size (water discharge).
    Blue (orange) lines represent outputs from the subglacial (subaerial) channel.
    Data are shown at $15$\unit{min} intervals.
    Arrows denote where variables peak in the subglacial channel before the subaerial one.
    Insets in f-h show the  peak melt event denoted by the shaded area in panels e--h, with an arbitrary y-axis.
    % Note that the x and y axes are different for \alpine{} and \icesheet{}.
  }
  \label{fig:model_outs}
\end{figure}

In the first experiment defined in Section~\ref{sect:imp}, model outputs exhibit different seasonal evolutions and peaks.
Water velocity and shear stress represent the potential for sediment mobilization, and width-integrated shear stress represents a proxy for the total sediment transport capacity across the channel bed (Figure~\ref{fig:model_outs}).
Additionally, the subglacial model outputs are generally larger than their subaerial equivalents, thus greater sediment transport capacity would be expected in subglacial channels compared to subaerial ones.

@@

Examining the relationship between water discharge and the model outputs, different relationships emerge between the subaerial and subglacial cases due to the hysteresis in channel size in the subglacial model (Equation~\ref{eq:dS_dt}).
Because there is no history dependence in  subaerial channels, each water discharge value in the subaerial channel results in a unique water velocity, shear stress, and width-integrated shear stress (Section~\ref{sect:sub_mode}).
This characteristic results in a perfect rank correlation between the variables (Figure~\ref{fig:Qw_vari}).

@@

The second numerical experiment aims to evaluate the role of channel shape and friction factors on the findings above and to examine the effect of variability in water discharge.
Across the range of channel shapes and friction factors  examined (Section\,\ref{sect:imp}), variability in water velocity and shear stress remains higher in the subglacial system compared to the subaerial one in both \alpine{} and \icesheet{}.
This suggests that there is greater intermittency and variability in sediment transport in subglacial systems compared to subaerial ones, especially in supply-limited systems where sediment could be exhausted or mobilized over diurnal discharge cycles.
Considering the width integrated shear stress, in \alpine{} the variability is comparable between the subaerial and subglacial systems, due to the highly variable water discharge, largely due to diurnal variations (Figure~\ref{fig:multi_run}).
However, in \icesheet{}, with lower water discharge fluctuations, the variability in width-integrated shear stress remains consistently higher in the subglacial system across the time aggregation periods.

Results from  both \alpine{} and \icesheet{} scenarios show that the variability in subglacial outputs decreases with aggregations longer than approximately 1--5 days (Figure~\ref{fig:multi_run}).
Yet, substantial differences in the variability between the model outputs for subglacial and subaerial cases persist in periods of up to $15$ days, longer than which seasonal variations might be conflated.
Note that across longer aggregation times, the timing of peak sediment transport conditions still occur in the natural environment. Yet they are not captured in model outputs(Figure~\ref{fig:model_outs}), despite their potential geomorphic importance.

@@

The different response to water discharge in the model outputs subglacial channels compared to subaerial ones (Figure~\ref{fig:Qw_vari}) may in fact cause even greater variations in sediment transport capacity in subglacial channels than presented above.
In sediment transport relationships that evaluate sediment discharge capacity, such as in \citeA{meyer1948} or \citeA{engelund1967}, shear stress is scaled to the power of $\frac{3}{2}$ or $\frac{5}{2}$, respectively.
The exponent greater than $1$ magnifies sediment discharge variability beyond the variable sediment transport parameters described above (Figure~\ref{fig:multi_run}; Table~\ref{tab:Qs}).
Variability increases further due to the application of the exponent occurs before integrating the sediment transport capacity across the channel width, as was done here in evaluating $w\tau$.

Sediment transport capacity in subglacial channels responds more rapidly to changes in water discharge compared to their subaerial counterparts \cite<Figures~\ref{fig:Qw_vari}~and~\ref{fig:multi_run}>{alley1997}.
In subglacial R\"othlisberger channels in steady state with water discharge, the width-integrated shear stress response to water discharge is $ w_g\,\tau_{g}\, \propto\,  Q^{\sim\frac{4}{5}}$ (Table\,\ref{tab:eqs1}).
However, subglacial channels rarely operate in a steady state \cite<e.g.>{gimbert2016}.
For sufficiently short periods, channel width and wetted area are fixed in the subglacial system (akin to pipe flow) so that the response of width-integrated shear stress is $w_g\, \tau_{g}\, \propto\,  Q^{2}$ (Table\,\ref{tab:eqs1}).
\ref{sect:scaling} provides a complete description of sediment transport capacity in different drainage conditions.
Note that in pipe flow conditions, maximum sediment transport would coincide with peak water discharge, unlike with the evolving channel size here where a disparate relationship with water discharge occurs (Section~\ref{sect:sub_mode}; Figure \ref{fig:model_outs} and \ref{fig:Qw_vari}). 
Similar adjustments in channel shape occur in subaerial channels \cite{phillips2016}, albeit at timescales longer than the subglacial channel's response time of hours to days.

Large catchments, or those with reduced water discharge variability, may experience yet less variability in sediment transport capacity, as the two hydrological forcings from \alpine{} and \icesheet{} show different variations in  sediment transport capacity.
Indeed, the more decoupled and sporadic relationship between model outputs and water discharge in \alpine{} results from the relatively larger discharge variations on daily to weekly timescales compared to \icesheet{}.
This occurs as the evolution of the R-channel's size is small compared to the variations in water discharge.
Such water discharge variability in alpine catchments may cause shear stress to approach the water discharge squared in assuming pipe-flow conditions \cite< \ref{sect:scaling}; c.f.>{alley1997}.
Conversely, the reduced variability in water discharge in \icesheet{} may result in the R-channel's size being closer to equilibrium concerning the variations in water discharge.
Due the smaller water discharge variability, a closer approximation of the hydraulics could be represented by pipe flow here, where the channel size remains fixed.
In this case, the exponent on water discharge for shear stress is likely less than $2$, but greater than $\frac{4}{5}$ occurring when the R-channel's size evolves in equilibrium with water discharge (see \ref{sect:scaling}).
As a result, there is a stronger between subglacial model outputs and water discharge in \icesheet{} compared to \alpine{} (Figure~\ref{fig:Qw_vari}) due to the channel size being in closer equilibrium with water discharge in the \icesheet{} case.

@@

Subglacial channels exhibit far weaker relationship between sediment transport capacity and water discharge than is the case in subaerial channels (Figure~\ref{fig:Qw_vari}).
In subaerial channels sediment transport variations in subaerial channels can often be attributed to hydro climatic conditions \cite{wolman1960}, which could be less straightforward in subglacial environments \cite{lane2019}. 
A glacier's sediment transport capacity is impacted by the ice thickness controlling the channel closure rate, and the glacier's surface slope, in addition to water discharge and sediment size \cite<Figure~\ref{fig:multi_run}, Section~\ref{sect:sub_mode}; >{rothlisberger1972,shreve1972,gimbert2016,stevens2022,walder1994}.
In a supply-limited state, a glacier's sediment discharge record also responds to  sediment availability and bedrock erosion from  water pressure variations and sliding  \cite<e.g.>{iverson2012,herman2015,delaney2023}.
This multitude of processes lies in contrast to many subaerial channels, where transport capacity typically responds to sediment size, channel shape, water discharge, and hydraulic gradient,
while the hydraulic gradient and channel shape can remain stable over years to millennia \cite<Section~\ref{sect:fluv}; e.g.>{tucker1997}.

@@

Greater variations in subglacial sediment transport capacity compared to subaerial ones could cause a supply-limited regime at many glaciers.
In subglacial systems, sediment's critical shear stress, or threshold at which sediment mobilization occurs, can be reached more frequently and across a range of water discharges, compared to subaerial systems (Figure~\ref{fig:Qw_vari}).
This result suggests sediment export in these catchments depend on water discharge variability, in addition to quantity \cite{lane2019}.
Sediment exhaustion through repeated crossing of the mobilization threshold may explain the stronger dependence of sediment discharge from the Greenland Ice Sheet on basal shear stress, a proxy for bedrock erosion, rather than glacier melt \cite{overeem2017}.
The repeated crossing of the mobilization threshold could also result in sediment discharge's strong dependence on sediment production from sliding in many cases \cite{herman2015,koppes2015}.

@@

Numerical models of subglacial sediment transport generally apply classical parameterizations of subglacial hydrology to evaluating sediment transport conditions \cite<e.g>{hewitt2019,beaud2018}.
In these R-channel models, realistic water pressure regimes can be achieved by changing the channels shape and friction factor, therefore  models are tuned with the friction factor and channel shape remains fixed as a semi-circle \cite<$\beta =\pi$; >{werder2010b}.
However, the role of channel width in the findings here suggests that more thorough parameterization of channel shape and channel width is needed in assessing subglacial sediment transport capacity compared to water pressure (Section\,\ref{sect:sub_mode}, \ref{sect:scaling}).

Water discharge measurements at the hourly timescale remain unavailable in many catchments, further limiting the ability of subglacial  sediment transport models to capture specific events.
The model output variability in subglacial channels approaches that of subaerial ones over $1-5$ days of  aggregation (Figure~\ref{fig:multi_run}), possibly resulting in parity in sediment transport capacity variations between subaerial and subglacial channels over these periods.
Note that although variability between the events could remain comparable, water discharge forcings aggregated over periods longer than days may omit key sediment discharge characteristics, such as the timing of peaks in sediment transport capacity from glaciers (Figure~\ref{fig:model_outs}).

@@

Implications of the presented findings include:
\begin{enumerate}
\item  evolving subglacial channel size causes  water discharge to have a changing impact on sediment transport capacity, complicating the relationship between sediment export and hydro-climatic conditions below glaciers;
  \vspace{.1cm}

\item the  variable relationship between water discharge and sediment transport capacity below glaciers creates larger variability in sediment transport capacity compared to river systems;
 \vspace{.1cm}

\item the uneven response of sediment transport capacity to water discharge variations shows a different response and timing to periods of peak discharge in subglacial channels compared to subaerial ones.
However, representing subglacial sediment transport capacity with either open-channel flow or pipe flow neglects differences in timing.
\end{enumerate}

This study calls for the careful analysis when examining the relationship between sediment transport capacity and hydro climatic conditions from glacierized catchments.
